---
title: "E-Commerce User Behavior Modeling through Classification, Regression and Predictive Analytics"
author: "Group 11_Project-WQD7004_PROGRAMMING FOR DATA SCIENCE"
date: "2025-06-12"
output: html_document
---

## R Markdown

Project Title "E-Commerce User Behavior Modeling through Classification, Regression and Predictive Analytics"

### Introduction

The “E-commerce Customer Behaviour and Purchase” dataset serves as a rich and structured representation of online-shopping-customer interactions within an E-commerce platform. The dataset from Kaggle is described as a synthetically generated dataset using the Python Faker library. Although it is not real-world data, the dataset is modelled to reflect or imitate realistic e-commerce behaviour patterns.

The dataset is generated to facilitate various types of data science and machine learning tasks related to online retail, such as customer churn prediction, market basket analysis, recommendation systems, and purchasing trend analysis etc. The dataset is well-structured in tabular form, where each row represents a unique transaction made by a customer, and each column captures a specific attribute related to the transaction, the customer, or their purchase behaviour.

The dataset spans multiple relevant aspects of customer purchasing behaviour. Key features include customer demographic information such as Customer ID, Customer Name, Age, and Gender, along with transactional data like Purchase Date, Product Category, Product Price, Quantity, and Total Purchase Amount. It also includes behavioural indicators such as Payment Method, returns and customer churn. These features collectively allow for in-depth analyses of consumer habits and provide a strong foundation for predictive modelling.

Two sets of datasets comprising 250000 records and 13 columns are provided. Given its use case in modelling customer behaviour over time, it can be assumed that it spans a large number of records to accurately simulate e-commerce activities across different customers and timeframes.

### Project Objectives

1. To identify factors influencing customer churn, helping businesses improve retention strategies.

2. To model user behavior using classification and regression techniques to predict important customer outcomes.


```{r }
# Reading the dataset
df <- read.csv("ecommerce_customer_data_large.csv", stringsAsFactors = FALSE)
```


```{r}
# Load necessary libraries
library(tidyverse)   # For data manipulation and visualization
library(caret)       # For machine learning functions
library(e1071)       # For SVM and other ML models
library(rpart)       # For decision trees
library(nnet)        # For neural networks
library(MASS)        # For logistic regression
library(ggplot2)     # For visualization
library(dplyr)       # For data wrangling```
library(lubridate)
```

```{r}
# View structure and summary (df)

# Structure (data types)
str(df)

# Quick overview
glimpse(df)

# Descriptive statistics
summary(df)

# Dimensions: Rows and Columns
dim(df)

# Column Names
colnames(df)

# Check for first few columns
head(df)
```

### Data Cleaning & Preprocessing

* Data Cleaning

Removing duplicates

Handling missing values

Fixing column names or formats

Converting data types (e.g., character to date)

Drop similar column (e.g., Customer_Age)

* Data Preprocessing

Encoding categorical variables into numeric formats

Feature scaling (normalization/standardization)

Creating new features

Binning or discretizing variables

```{r}
# Check how many duplicate rows exist
sum(duplicated(df))
```

```{r}
# Check for missing values
colSums(is.na(df))
```

```{r}
# Check rows with missing values in 'Returns'
missing_rows <- df[is.na(df$Returns), ]

 # View the first few rows with missing values
head(missing_rows)
```

```{r}
# Replace all dots with underscores in column names
colnames(df) <- gsub("\\.", "_", colnames(df))

# Check the column name after change
colnames(df)
```

```{r}
# Convert data types
df$Purchase_Date <- as.Date(df$Purchase_Date, format = "%Y-%m-%d")
df$Returns <- as.integer(df$Returns)

# Create new features
# Extract components
df$Purchase_Day_Name <- weekdays(df$Purchase_Date)
df$Purchase_Month <- month(df$Purchase_Date)
df$Purchase_Month_Name <- month(df$Purchase_Date, label = TRUE, abbr = FALSE)
df$Purchase_Year <- format(df$Purchase_Date, "%Y")
head(df)
```

```{r}
# Check for total number of numerical features and categorical features in the dataset
num_features <- sum(sapply(df, is.numeric))
cat_features <- sum(sapply(df, function(x) is.factor(x) || is.character(x)))

cat("Numerical features:", num_features, "\n")
cat("Categorical features:", cat_features, "\n")
```

```{r}
# Show the names of the numerical and categorical features
numerical_features <- names(df)[sapply(df, is.numeric)]
categorical_features <- names(df)[sapply(df, function(x) is.factor(x) || is.character(x))]

cat("Numerical Features:\n")
print(numerical_features)

cat("\nCategorical Features:\n")
print(categorical_features)
```

```{r}
# Count the total number of churn and return
# Total Churned
sum(df$Churn == 1)

# Total Returned
sum(df$Returns == 1)
```

Notice that for the sum of returned, the result showed , which means there is NA in the dataset for 'Returns'

```{r}
# Count the total number of return (only with value, set na.rm to TRUE)
total_returns <- sum(df$Returns == 1, na.rm = TRUE)
print(total_returns)
```

```{r}
# Assuming the NA (missing value) is no return, we fill up with '0'
df$Returns[is.na(df$Returns)] <- 0
total_returns <- sum(df$Returns == 1)
print(total_returns)
```

```{r}
# Check whether there is still NA in the dataset
sum(is.na(df$returns))
```

```{r}
# Count the total number of churn and return
# Total Churned
sum(df$Churn == 1)

# Total Returned
sum(df$Returns == 1)
```

There is no NA value in the dataset anymore, which means we successfully fill up missing value with '0'
There is a total of 101476 of returns.


### DATA ANALYSIS - EDA

Total Revenue and Average Revenue
```{r}
# Calculate for total revenue and average revenue

# Total Revenue
total_revenue <- sum(df$Total_Purchase_Amount, na.rm = TRUE)
print(total_revenue)
# Average Revenue
average_revenue <- mean(df$Total_Purchase_Amount, na.rm = TRUE)
print(average_revenue)
```

Top 5 Spending Customers
```{r}
# Show the top 5 spending customers

top_customers <- df %>%
  group_by(Customer_ID, Customer_Name) %>%
  summarise(Total_Spending = sum(Total_Purchase_Amount, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Total_Spending)) %>%
  slice_head(n = 5)

top_customers
```

```{r}
# Plot bar graph
ggplot(top_customers, aes(x = reorder(Customer_Name, -Total_Spending), y = Total_Spending, fill = Customer_Name)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 5 Highest Spending Customers",
       x = "Customer Name",
       y = "Total Spending") +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_text(aes(label = round(Total_Spending, 2)), vjust = -0.3)
```
Busiest Sales Day of the Week
```{r}
# Determine the Busiest Sales Day of the week
sales_by_day <- df %>%
  group_by(Purchase_Day_Name) %>%
  summarise(Total_Sales = sum(Total_Purchase_Amount, na.rm = TRUE)) %>%
  arrange(desc(Total_Sales))

# Show result
print(sales_by_day)
```

```{r}
# Reorder factor based on Total_Sales (in descending order)
sales_by_day$Purchase_Day_Name <- factor(sales_by_day$Purchase_Day_Name, levels = sales_by_day$Purchase_Day_Name)

# Plot bar graph
ggplot(sales_by_day, aes(x = Purchase_Day_Name, y = Total_Sales, fill = Purchase_Day_Name)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Sales by Day of the Week", x = "Day", y = "Total Sales") +
  theme_minimal() +
  theme(legend.position = "none")
```

Product Category
```{r}
# List out all the product category
df %>%
  distinct(Product_Category)
```

```{r}
# Count products per category
category_counts <- df %>%
  group_by(Product_Category) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100,
         Label = paste0(Product_Category, " (", round(Percentage, 1), "%)"))

# Plot pie chart
ggplot(category_counts, aes(x = "", y = Count, fill = Product_Category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5), size = 4) +
  labs(title = "Product Category Distribution") +
  theme_void()
```

Payment Type
```{r}
# List out all the payment type
df %>%
  distinct(Payment_Method)
```

```{r}
# Count number of transactions per payment method
payment_counts <- df %>%
  group_by(Payment_Method) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

# Plot bar graph
ggplot(payment_counts, aes(x = reorder(Payment_Method, -Count), y = Count, fill = Payment_Method)) +
  geom_bar(stat = "identity") +
  labs(title = "Transaction Count by Payment Method",
       x = "Payment Method",
       y = "Number of Transactions") +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_text(aes(label = Count), vjust = -0.3)
```

Univariate Analysis (Numerical Variables)
Univariate Analysis

Numerical Variables (Customer_Age, Product_Price, Total_Purchase_Amount)
• Understand distribution, outliers, and spread

hist(): Used to visualize the distribution of a numerical feature.

boxplot(): Used to detect outliers and visualize spread.

summary(): Provides a quick overview of key statistics like mean, median, and quartiles.

```{r}
# Customer Age Histogram
hist(df$Age,
     main = "Histogram of Customer Age",
     xlab = "Customer Age",
     col = "lightblue",
     border = "black",
     breaks = 15)  # Adjust number of bins with breaks

# Product Price Histogram
hist(df$Product_Price,
     main = "Histogram of Product Price",
     xlab = "Product Price",
     col = "lightgreen",
     border = "black",
     breaks = 20)

# Total Purchase Amount Histogram
hist(df$Total_Purchase_Amount,
     main = "Histogram of Total Purchase Amount",
     xlab = "Total Purchase Amount",
     col = "lightcoral",
     border = "black",
     breaks = 25)
```

```{r}
# Customer Age Boxplot
boxplot(df$Age,
        main = "Boxplot of Customer Age",
        ylab = "Customer Age",
        col = "lightblue",
        border = "blue")

# Product Price Boxplot
boxplot(df$Product_Price,
        main = "Boxplot of Product Price",
        ylab = "Product Price",
        col = "lightgreen",
        border = "green")

# Total Purchase Amount Boxplot
boxplot(df$Total_Purchase_Amount,
        main = "Boxplot of Total Purchase Amount",
        ylab = "Total Purchase Amount",
        col = "lightcoral",
        border = "red")
```

```{r}
# Summary for Customer Age
summary(df$Age)

# Summary for Product Price
summary(df$Product_Price)

# Summary for Total Purchase Amount
summary(df$Total_Purchase_Amount)
```

Univariate Analysis (Categorical Variables)
Categorical Variables (Gender, Product_Category, Payment_Method, Returns, Churn)
• See frequency of categories

table(): Generates a frequency table for categorical data, showing how many times each category appears.

barplot(): Visualizes categorical data using bar charts.

prop.table(): Converts the frequency table to proportions (useful for relative frequency analysis).

```{r}
# Frequency Table for Gender
table(df$Gender)

# Frequency Table for Product Category
table(df$Product_Category)

# Frequency Table for Payment Method
table(df$Payment_Method)

# Frequency Table for Returns
table(df$Returns)

# Frequency Table for Churn
table(df$Churn)
```

```{r}
# Bar Plot for Gender
gender_counts <- table(df$Gender)
barplot(gender_counts,
        main = "Gender Distribution",
        col = c("lightblue", "lightgreen"),
        xlab = "Gender",
        ylab = "Count",
        names.arg = names(gender_counts))

# Bar Plot for Product Category
product_category_counts <- table(df$Product_Category)
barplot(product_category_counts,
        main = "Product Category Distribution",
        col = rainbow(length(product_category_counts)),
        xlab = "Product Category",
        ylab = "Count",
        las = 2)  # Rotate axis labels for better readability

# Bar Plot for Payment Method
payment_method_counts <- table(df$Payment_Method)
barplot(payment_method_counts,
        main = "Payment Method Distribution",
        col = c("lightblue", "lightcoral", "lightgreen"),
        xlab = "Payment Method",
        ylab = "Count",
        names.arg = names(payment_method_counts))
# Bar Plot for Returns
returns_counts <- table(df$Returns)
barplot(returns_counts,
        main = "Returns Distribution",
        col = c("lightblue", "lightpink"),
        xlab = "Returns",
        ylab = "Count",
        names.arg = names(returns_counts))

# Bar Plot for Churn
churn_counts <- table(df$Churn)
barplot(churn_counts,
        main = "Churn Distribution",
        col = c("lightgreen", "lightcoral"),
        xlab = "Churn",
        ylab = "Count",
        names.arg = names(churn_counts))
```

```{r}
# Bar Plot with Proportions for Gender
gender_proportions <- prop.table(table(df$Gender))
barplot(gender_proportions,
        main = "Gender Distribution (Proportions)",
        col = c("lightblue", "lightgreen"),
        xlab = "Gender",
        ylab = "Proportion",
        names.arg = names(gender_proportions))



# Bar Plot with Proportions for Product Category
product_category_proportions <- prop.table(table(df$Product_Category))
barplot(product_category_proportions,
        main = "Product Category Distribution (Proportions)",
        col = c("lightblue", "lightgreen"),
        xlab = "Product Category",
        ylab = "Proportion",
        names.arg = names(product_category_proportions))


# Bar Plot with Proportions for Payment Method
payment_method_proportions <- prop.table(table(df$Payment_Method))
barplot(payment_method_proportions,
        main = "Payment Method Distribution (Proportions)",
        col = c("lightblue", "lightgreen"),
        xlab = "Payment Method",
        ylab = "Proportion",
        names.arg = names(payment_method_proportions))

# Bar Plot with Proportions for Returns
returns_proportions <- prop.table(table(df$Returns))
barplot(returns_proportions,
        main = "Payment Method Distribution (Proportions)",
        col = c("lightblue", "lightgreen"),
        xlab = "Returns",
        ylab = "Proportion",
        names.arg = names(returns_proportions))


# Bar Plot with Proportions for Churn
churn_proportions <- prop.table(table(df$Churn))
barplot(churn_proportions,
        main = "Churn Distribution (Proportions)",
        col = c("lightgreen", "lightcoral"),
        xlab = "Churn",
        ylab = "Proportion",
        names.arg = names(churn_proportions))
```

Bivariate Analysis (Churn VS Features)

```{r}
# Boxplot to compare Total Purchase Amount by Churn status
boxplot(Total_Purchase_Amount ~ Churn,
        data = df,
        main = "Comparison of Total Purchase Amount by Churn Status",
        xlab = "Churn",
        ylab = "Total Purchase Amount",
        col = c("lightblue", "lightcoral"),
        names = c("Non-Churned", "Churned"))

# Boxplot for Customer Age vs. Churn status
boxplot(Age ~ Churn,
        data = df,
        main = "Comparison of Customer Age by Churn Status",
        xlab = "Churn",
        ylab = "Customer Age",
        col = c("lightblue", "lightcoral"),
        names = c("Non-Churned", "Churned"))

# Boxplot for Product Price vs. Churn status
boxplot(Product_Price ~ Churn,
        data = df,
        main = "Comparison of Product Price by Churn Status",
        xlab = "Churn",
        ylab = "Product Price",
        col = c("lightblue", "lightcoral"),
        names = c("Non-Churned", "Churned"))
```

```{r}
library(RColorBrewer)

# Create a bar plot for Gender vs Churn
gender_churn_table <- table(df$Gender, df$Churn)
barplot(gender_churn_table,
        beside = TRUE,
        col = c("lightblue", "lightcoral"),
        legend = rownames(gender_churn_table),
        main = "Gender vs Churn",
        xlab = "Churn Status",
        ylab = "Count",
        names.arg = c("Non-Churned", "Churned"))

# Create a bar plot for Payment Method vs Churn
payment_method_churn_table <- table(df$Payment_Method, df$Churn)
barplot(payment_method_churn_table,
        beside = TRUE,
        col = c("lightblue", "lightcoral", "lightyellow"),
        legend = rownames(payment_method_churn_table),
        main = "Payment Method vs Churn",
        xlab = "Payment Method Status",
        ylab = "Count",
        names.arg = c("Non-Churned", "Churned"))

# Create a bar plot for Product Category vs Churn
product_category_churn_table <- table(df$Product_Category, df$Churn)
barplot(product_category_churn_table,
        beside = TRUE,
        col = c("lightblue", "lightcoral", "lightyellow", "lightgreen"),
        legend = rownames(product_category_churn_table),
        main = "Product Category vs Churn",
        xlab = "Churn Status",
        ylab = "Count",
        names.arg = c("Non-Churned", "Churned"))

# Create a bar plot for Purchase Month Name vs Churn
month_colors <- brewer.pal(12, "Set3")
purchase_month_name_churn_table <- table(df$Purchase_Month_Name, df$Churn)
barplot(purchase_month_name_churn_table,
        beside = TRUE,
        col = month_colors,
        legend = rownames(purchase_month_name_churn_table),
        main = "Purchase Month Name vs Churn",
        xlab = "Churn Status",
        ylab = "Count",
        names.arg = c("Non-Churned", "Churned"))
```

Bivariate Analysis (Return VS Features)
```{r}
# Boxplot to compare Total Purchase Amount by Return status
boxplot(Total_Purchase_Amount ~ Returns,
        data = df,
        main = "Comparison of Total Purchase Amount by Return Status",
        xlab = "Return",
        ylab = "Total Purchase Amount",
        col = c("lightblue", "lightcoral"),
        names = c("Non-Returned", "Returned"))

# Boxplot for Customer Age vs. Return status
boxplot(Age ~ Returns,
        data = df,
        main = "Comparison of Customer Age by Return Status",
        xlab = "Return",
        ylab = "Customer Age",
        col = c("lightblue", "lightcoral"),
        names = c("Non-Returned", "Returned"))

# Boxplot for Product Price vs. Return status
boxplot(Product_Price ~ Returns,
        data = df,
        main = "Comparison of Product Price by Return Status",
        xlab = "Return",
        ylab = "Product Price",
        col = c("lightblue", "lightcoral"),
        nnames = c("Non-Returned", "Returned"))
```

```{r}
# Contingency table for Gender vs Returns
table(df$Gender, df$Returns)

# Contingency table for Payment Method vs Returns
table(df$Payment_Method, df$Returns)

# Contingency table for Product Category vs Returns
table(df$Product_Category, df$Returns)

# Contingency table for Purchase Month Name vs Returns
table(df$Purchase_Month_Name, df$Returns)
```

```{r}
# Create a bar plot for Gender vs Returns
gender_returns_table <- table(df$Gender, df$Returns)
barplot(gender_returns_table,
        beside = TRUE,
        col = c("lightblue", "lightcoral"),
        legend = rownames(gender_returns_table),
        main = "Gender vs Returns",
        xlab = "Returns Status",
        ylab = "Count",
        names.arg = c("Non-Returned", "Returned"))

# Create a bar plot for Payment Method vs Returns
payment_method_returns_table <- table(df$Payment_Method, df$Returns)
barplot(payment_method_returns_table,
        beside = TRUE,
        col = c("lightblue", "lightcoral", "lightyellow"),
        legend = rownames(payment_method_returns_table),
        main = "Payment Method vs Returns",
        xlab = "Returns Status",
        ylab = "Count",
        names.arg = c("Non-Returned", "Returned"))
# Create a bar plot for Product Category vs Returns
product_category_returns_table <- table(df$Product_Category, df$Returns)
barplot(product_category_returns_table,
        beside = TRUE,
        col = c("lightblue", "lightcoral", "lightyellow", "lightgreen"),
        legend = rownames(product_category_returns_table),
        main = "Product Category vs Returns",
        xlab = "Returns Status",
        ylab = "Count",
        names.arg = c("Non-Returned", "Returned"))

# Create a bar plot for Purchase Month Name vs Returns
month_colors <- brewer.pal(12, "Set3")
purchase_month_name_returns_table <- table(df$Purchase_Month_Name, df$Returns)
barplot(purchase_month_name_returns_table,
        beside = TRUE,
        col = month_colors,
        legend = rownames(purchase_month_name_returns_table),
        main = "Purchase Month Name vs Returns",
        xlab = "Returns Status",
        ylab = "Count",
        names.arg = c("Non-Returned", "Returned"))
```

Return Rate
Return Rate (by product category, gender, purchase month name, payment method)
```{r}
# Return rate by product category
return_rate <- df %>%
  group_by(Product_Category) %>%
  summarise(Returns = sum(Returns == "1", na.rm = TRUE),
            Total_Orders = n(),
            Return_Rate = Returns / Total_Orders)

print(return_rate)
```

```{r}
# Return rate by gender
return_rate_gender <- df %>%
  group_by(Gender) %>%
  summarise(Returns = sum(Returns == "1", na.rm = TRUE),
            Total_Orders = n(),
            Return_Rate = Returns / Total_Orders)

print(return_rate_gender)
```

```{r}
ggplot(return_rate_gender, aes(x = Gender, y = Return_Rate, fill = Gender)) +
  geom_bar(stat = "identity") +
  labs(title = "Return Rate by Gender", x = "Gender", y = "Return Rate") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format())
```

```{r}
# Return rate by purchase month name
return_rate_month <- df %>%
  group_by(Purchase_Month_Name) %>%
  summarise(Returns = sum(Returns == "1", na.rm = TRUE),
            Total_Orders = n(),
            Return_Rate = Returns / Total_Orders)

print(return_rate_month)
```


```{r}
return_rate_month$Purchase_Month_Name <- factor(return_rate_month$Purchase_Month_Name, levels = month.name)

ggplot(return_rate_month, aes(x = Purchase_Month_Name, y = Return_Rate, fill = Purchase_Month_Name)) +
  geom_bar(stat = "identity") +
  labs(title = "Return Rate by Month", x = "Month", y = "Return Rate") +
  theme_minimal() +
  scale_fill_manual(values = brewer.pal(12, "Set3")) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Return rate by payment method
return_rate_payment <- df %>%
  group_by(Payment_Method) %>%
  summarise(Returns = sum(Returns == "1", na.rm = TRUE),
            Total_Orders = n(),
            Return_Rate = Returns / Total_Orders)

print(return_rate_payment)
```

```{r}
ggplot(return_rate_payment, aes(x = Payment_Method, y = Return_Rate, fill = Payment_Method)) +
  geom_bar(stat = "identity") +
  labs(title = "Return Rate by Payment Method", x = "Payment Method", y = "Return Rate") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Churn Rate
Churn Rate (by product category, gender, purchase month name, payment method)

```{r}
# Churn rate by product category
churn_rate_category <- df %>%
  group_by(Product_Category) %>%
  summarise(Churned = sum(Churn == "1", na.rm = TRUE),
            Total_Customers = n(),
            Churn_Rate = Churned / Total_Customers)

print(churn_rate_category)
```

```{r}
ggplot(churn_rate_category, aes(x = reorder(Product_Category, -Churn_Rate), y = Churn_Rate, fill = Product_Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Churn Rate by Product Category", x = "Product Category", y = "Churn Rate") +
  theme_minimal() +
  scale_fill_manual(values = brewer.pal(n = min(12, length(unique(churn_rate_category$Product_Category))), name = "Set3")) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

```{r}
# Churn rate by gender
churn_rate_gender <- df %>%
  group_by(Gender) %>%
  summarise(Churned = sum(Churn == "1", na.rm = TRUE),
            Total_Customers = n(),
            Churn_Rate = Churned / Total_Customers)

print(churn_rate_gender)
```

```{r}
ggplot(churn_rate_gender, aes(x = Gender, y = Churn_Rate, fill = Gender)) +
  geom_bar(stat = "identity") +
  labs(title = "Churn Rate by Gender", x = "Gender", y = "Churn Rate") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format())
```

```{r}
# Churn rate by purchase month name
churn_rate_month <- df %>%
  group_by(Purchase_Month_Name) %>%
  summarise(Churned = sum(Churn == "1", na.rm = TRUE),
            Total_Customers = n(),
            Churn_Rate = Churned / Total_Customers)

# Sort month names properly
churn_rate_month$Purchase_Month_Name <- factor(churn_rate_month$Purchase_Month_Name, levels = month.name)

print(churn_rate_month)
```

```{r}
ggplot(churn_rate_month, aes(x = Purchase_Month_Name, y = Churn_Rate, fill = Purchase_Month_Name)) +
  geom_bar(stat = "identity") +
  labs(title = "Churn Rate by Month", x = "Month", y = "Churn Rate") +
  theme_minimal() +
  scale_fill_manual(values = brewer.pal(12, "Set3")) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Churn rate by payment method
churn_rate_payment <- df %>%
  group_by(Payment_Method) %>%
  summarise(Churned = sum(Churn == "1", na.rm = TRUE),
            Total_Customers = n(),
            Churn_Rate = Churned / Total_Customers)

print(churn_rate_payment)
```

```{r}
ggplot(churn_rate_payment, aes(x = Payment_Method, y = Churn_Rate, fill = Payment_Method)) +
  geom_bar(stat = "identity") +
  labs(title = "Churn Rate by Payment Method", x = "Payment Method", y = "Churn Rate") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```


### Data Modelling

#### 1. Classification (Churn Prediction)

```{r}
# Required libraries
library(caret) # Classification and Regression Training
library(e1071) # Tools for SVM, Naive Bayes, and more
library(randomForest) # Random Forest Algorithm
library(xgboost) # Extreme Gradient Boosting
library(MLmetrics) # Evaluation Metrics for ML Models
library(lubridate)
library(dplyr)
```

```{r}
# Load and Prepare Data
df <- read.csv("ecommerce_customer_data_large.csv", stringsAsFactors = FALSE)
colnames(df) <- gsub("\\.", "_", colnames(df))
df$Purchase_Date <- as.Date(df$Purchase_Date, format = "%Y-%m-%d")
df$Returns <- as.integer(df$Returns)
df <- df %>% dplyr::select(-Customer_Age)
```

```{r}
# Convert relevant columns to factors
df$Gender <- as.factor(df$Gender)
df$Payment_Method <- as.factor(df$Payment_Method)
df$Product_Category <- as.factor(df$Product_Category)
df$Churn <- as.factor(df$Churn) # Target
```

```{r}
# Train-test split
set.seed(42)
trainIndex <- createDataPartition(df$Churn, p = 0.80, list = FALSE)
train <- df[trainIndex, ]
test <- df[-trainIndex, ]

# Select relevant features
selected_vars <- c("Churn", "Product_Category", "Product_Price", "Quantity",
                   "Total_Purchase_Amount", "Payment_Method", "Age", "Gender")
train <- train[, selected_vars]
test <- test[, selected_vars]
```


Customer Churn (Classification)

Model to be use:
1. Logistic Regression
2. Random Forest
3. XGBoost


##### 1. Logistic Regression (Churn Prediction)

```{r}
# Fit model
model_log <- glm(
  Churn ~ ., data = train, family = "binomial"
)

# Predict on test data
pred_prob_log <- predict(model_log, newdata = test, type = "response")
pred_log <- factor(ifelse(pred_prob_log > 0.5, "1", "0"), levels = levels(test$Churn))

# Evaluation
conf_mat_log <- confusionMatrix(pred_log, test$Churn, positive = "1")
print(conf_mat_log)

# Metrics
accuracy_log <- conf_mat_log$overall["Accuracy"]
precision_log <- conf_mat_log$byClass["Pos Pred Value"]
recall_log <- conf_mat_log$byClass["Sensitivity"]
f1_log <- 2 * ((precision_log * recall_log) / (precision_log + recall_log))

cat("\n--- Logistic Regression Metrics ---\n")
cat("Accuracy:", accuracy_log, "\n")
cat("Precision:", precision_log, "\n")
cat("Recall:", recall_log, "\n")
cat("F1 Score:", f1_log, "\n")
```


##### 2. Random Forest (Churn Prediction)

```{r}
set.seed(42)
model_rf <- randomForest(Churn ~ ., data = train, ntree = 100)
pred_rf <- predict(model_rf, test)

# Evaluation
conf_mat_rf <- confusionMatrix(pred_rf, test$Churn, positive = "1")
print(conf_mat_rf)

# Metrics
accuracy_rf <- conf_mat_rf$overall["Accuracy"]
precision_rf <- conf_mat_rf$byClass["Pos Pred Value"]
recall_rf <- conf_mat_rf$byClass["Sensitivity"]
f1_rf <- 2 * ((precision_rf * recall_rf) / (precision_rf + recall_rf))

cat("\n--- Random Forest Metrics ---\n")
cat("Accuracy:", accuracy_rf, "\n")
cat("Precision:", precision_rf, "\n")
cat("Recall:", recall_rf, "\n")
cat("F1 Score:", f1_rf, "\n")
```


##### 3. XGBoost (Churn Prediction)

```{r}
# Prepare design matrices
train_matrix <- model.matrix(Churn ~ . -1, data = train)
test_matrix <- model.matrix(Churn ~ . -1, data = test)

# Prepare binary labels
train_label <- ifelse(train$Churn == "1", 1, 0)
test_label <- ifelse(test$Churn == "1", 1, 0)

# DMatrix objects
dtrain <- xgb.DMatrix(data = train_matrix, label = train_label)
dtest <- xgb.DMatrix(data = test_matrix, label = test_label)

# Set parameters and train
params <- list(
  objective = "binary:logistic",
  eval_metric = "logloss"
)

set.seed(42)
model_xgb <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 100,
  verbose = 0
)

# Predict
pred_prob_xgb <- predict(model_xgb, dtest)
pred_xgb <- factor(ifelse(pred_prob_xgb > 0.5, "1", "0"), levels = levels(test$Churn))

# Evaluation
conf_mat_xgb <- confusionMatrix(pred_xgb, test$Churn, positive = "1")
print(conf_mat_xgb)

# Metrics
accuracy_xgb <- conf_mat_xgb$overall["Accuracy"]
precision_xgb <- conf_mat_xgb$byClass["Pos Pred Value"]
recall_xgb <- conf_mat_xgb$byClass["Sensitivity"]
f1_xgb <- 2 * ((precision_xgb * recall_xgb) / (precision_xgb + recall_xgb))

cat("\n--- XGBoost Metrics ---\n")
cat("Accuracy:", accuracy_xgb, "\n")
cat("Precision:", precision_xgb, "\n")
cat("Recall:", recall_xgb, "\n")
cat("F1 Score:", f1_xgb, "\n")
```


##### Summary Table

```{r}
# Extract and compute metrics after each model evaluation

# Logistic Regression
log_accuracy <- as.numeric(conf_mat_log$overall["Accuracy"])
log_precision <- as.numeric(conf_mat_log$byClass["Pos Pred Value"])
log_recall <- as.numeric(conf_mat_log$byClass["Sensitivity"])
log_f1 <- 2 * ((log_precision * log_recall) / (log_precision + log_recall))

# Random Forest
rf_accuracy <- as.numeric(conf_mat_rf$overall["Accuracy"])
rf_precision <- as.numeric(conf_mat_rf$byClass["Pos Pred Value"])
rf_recall <- as.numeric(conf_mat_rf$byClass["Sensitivity"])
rf_f1 <- 2 * ((rf_precision * rf_recall) / (rf_precision + rf_recall))

# XGBoost
xgb_accuracy <- as.numeric(conf_mat_xgb$overall["Accuracy"])
xgb_precision <- as.numeric(conf_mat_xgb$byClass["Pos Pred Value"])
xgb_recall <- as.numeric(conf_mat_xgb$byClass["Sensitivity"])
xgb_f1 <- 2 * ((xgb_precision * xgb_recall) / (xgb_precision + xgb_recall))

# Create summary table
summary_table <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "XGBoost"),
  Accuracy = c(log_accuracy, rf_accuracy, xgb_accuracy),
  Precision = c(log_precision, rf_precision, xgb_precision),
  Recall = c(log_recall, rf_recall, xgb_recall),
  F1_Score = c(log_f1, rf_f1, xgb_f1),
  stringsAsFactors = FALSE
)

# Round numeric columns
summary_table[, 2:5] <- round(summary_table[, 2:5], 6)

# View summary
print(summary_table)
xgb_accuracy <- conf_mat_xgb$overall["Accuracy"]
xgb_precision <- conf_mat_xgb$byClass["Pos Pred Value"]
xgb_recall <- conf_mat_xgb$byClass["Sensitivity"]
xgb_f1 <- 2 * ((xgb_precision * xgb_recall) / (xgb_precision + xgb_recall))
```

#### SMOTE
```{r}
# Read the file
df <- read.csv("ecommerce_customer_data_large.csv")
head(df)

# Clean column names
colnames(df) <- gsub("\\.", "_", colnames(df))


# Convert data types
df$Purchase_Date <- as.Date(df$Purchase_Date, format = "%Y-%m-%d")
df$Returns[is.na(df$Returns)] <- 0
df$Returns <- as.integer(df$Returns)

# Drop redundant column
df <- df %>% dplyr::select(-Customer_Age)

# Extract date-related features
df$Purchase_Day_Name <- weekdays(df$Purchase_Date)
df$Purchase_Month <- month(df$Purchase_Date)
df$Purchase_Month_Name <- month(df$Purchase_Date, label = TRUE, abbr = FALSE)
df$Purchase_Year <- format(df$Purchase_Date, "%Y")

# Drop non-numeric or unnecessary columns for modeling
df_numeric <- df %>% dplyr::select(-Purchase_Date)

# Convert all character/factor variables to numeric
df_numeric[] <- lapply(df_numeric, function(x) {
  if (is.factor(x) || is.character(x)) {
    return(as.numeric(as.factor(x)))
  } else {
    return(x)
  }
})
```

```{r}
#load SMOTE library
library(smotefamily)
```

```{r}
# Apply SMOTE
set.seed(42)
smote_result <- SMOTE(
  X = df_numeric[, -which(names(df_numeric) == "Churn")],
  target = df_numeric$Churn,
  K = 5
)

df_smote <- smote_result$data
df_smote$Churn <- factor(df_smote$class, levels = c(0, 1))
df_smote$class <- NULL

# Split data
set.seed(42)
trainIndex <- createDataPartition(df_smote$Churn, p = 0.8, list = FALSE)
train_smote <- df_smote[trainIndex, ]
test_smote <- df_smote[-trainIndex, ]
```


##### 1. Logistic Regression (with SMOTE)

```{r}
model_log <- glm(Churn ~ ., data = train_smote, family = "binomial")
pred_prob_log <- predict(model_log, newdata = test_smote, type = "response")
pred_log <- factor(ifelse(pred_prob_log > 0.5, "1", "0"), levels = c("0", "1"))
test_smote$Churn <- factor(test_smote$Churn, levels = c("0", "1"))
conf_log <- confusionMatrix(pred_log, test_smote$Churn, positive = "1")

# Metrics
f1_log <- 2 * (conf_log$byClass["Pos Pred Value"] * conf_log$byClass["Sensitivity"]) /
  (conf_log$byClass["Pos Pred Value"] + conf_log$byClass["Sensitivity"])
```

##### 2. Random Forest (with SMOTE)

```{r}
model_rf <- randomForest(Churn ~ ., data = train_smote, ntree = 100)
pred_rf <- predict(model_rf, test_smote)
pred_rf <- factor(pred_rf, levels = c("0", "1"))
conf_rf <- confusionMatrix(pred_rf, test_smote$Churn, positive = "1")
f1_rf <- 2 * (conf_rf$byClass["Pos Pred Value"] * conf_rf$byClass["Sensitivity"]) /
  (conf_rf$byClass["Pos Pred Value"] + conf_rf$byClass["Sensitivity"])
```

##### 3. XGBoost (with SMOTE)

```{r}
train_matrix <- model.matrix(Churn ~ . -1, data = train_smote)
test_matrix <- model.matrix(Churn ~ . -1, data = test_smote)
train_label <- as.numeric(as.character(train_smote$Churn))
test_label <- as.numeric(as.character(test_smote$Churn))
dtrain <- xgb.DMatrix(data = train_matrix, label = train_label)
dtest <- xgb.DMatrix(data = test_matrix, label = test_label)

params <- list(
  objective = "binary:logistic",
  eval_metric = "logloss"
)

model_xgb <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 100,
  verbose = 0
)

pred_prob_xgb <- predict(model_xgb, dtest)
pred_xgb <- factor(ifelse(pred_prob_xgb > 0.5, "1", "0"), levels = c("0", "1"))
conf_xgb <- confusionMatrix(pred_xgb, test_smote$Churn, positive = "1")
f1_xgb <- 2 * (conf_xgb$byClass["Pos Pred Value"] * conf_xgb$byClass["Sensitivity"]) /
  (conf_xgb$byClass["Pos Pred Value"] + conf_xgb$byClass["Sensitivity"])
```


##### Summary
```{r}
cat("\n--- Logistic Regression Metrics ---\n")
print(conf_log)
cat("F1 Score:", f1_log, "\n")

cat("\n--- Random Forest Metrics ---\n")
print(conf_rf)
cat("F1 Score:", f1_rf, "\n")

cat("\n--- XGBoost Metrics ---\n")
print(conf_xgb)
cat("F1 Score:", f1_xgb, "\n")
```


#### 2. Regression

Finding highest correlation numeric features and categorical features (annova hypo test)
```{r}
library(corrplot)
```

```{r}
# Correlation matrix for numeric variables (all features do not show any correlation other than age)
num_vars <- df %>% dplyr::select(where(is.numeric))
cor_matrix <- cor(num_vars, use = "complete.obs")
corrplot(cor_matrix, method = "color", tl.cex = 0.8)

# Scatter plots vs. Total Purchase Amount
ggplot(df, aes(x = Product_Price, y = Total_Purchase_Amount)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm")

ggplot(df, aes(x = Quantity, y = Total_Purchase_Amount)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm")

ggplot(df, aes(x = Age, y = Total_Purchase_Amount)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm")

```

```{r}
# Correlations with the target variable
target_corr <- cor_matrix[, "Total_Purchase_Amount"]
# Sort in decreasing order
sorted_corr <- sort(abs(target_corr), decreasing = TRUE)
print(sorted_corr)
# highest correlation - age (0.056), Purchase_month (0.0016)
```

```{r}
# to test the significant imapct of categorical features
#(using annova) (paymentMethod show highest impact, lowest p value)

# Convert categorical variables to factors
df$Product_Category <- as.factor(df$Product_Category)
df$Payment_Method <- as.factor(df$Payment_Method)
df$Gender <- as.factor(df$Gender)

# Convert Purchase_Date to date format and extract useful features
df$Purchase_Date <- as.POSIXct(df$Purchase_Date)
df$Purchase_Year <- as.factor(format(df$Purchase_Date, "%Y"))
df$Purchase_Month <- as.factor(format(df$Purchase_Date, "%m"))
df$Purchase_Day <- as.factor(format(df$Purchase_Date, "%d"))
df$Purchase_Weekday <- as.factor(weekdays(df$Purchase_Date))

# For each categorical variable, perform ANOVA with Total_Purchase_Amount
anova_results <- aov(Total_Purchase_Amount ~ Product_Category, data = df)
summary(anova_results)

anova_results <- aov(Total_Purchase_Amount ~ Payment_Method, data = df)
summary(anova_results)

anova_results <- aov(Total_Purchase_Amount ~ Gender, data = df)
summary(anova_results)
# best pvalue - product_category and payment method (0.0782 and 0.0029 respectively)
```
Keep the most statistically relevant feautures

1. Age
2. Product_Category
3. Payment_Method
```{r}
colnames(df)
```

```{r}
# Read the file
df_model <- df %>%
  dplyr::select(Total_Purchase_Amount, Age, Product_Category, Payment_Method) %>%
  na.omit()
df_model$Product_Category <- as.factor(df_model$Product_Category)
df_model$Payment_Method <- as.factor(df_model$Payment_Method)

# Create polynomial terms (Age^2)  for polynomial regression
#df_model$Customer_Age2 <- df_model$Customer_Age^2
df_model$Age2 <- df_model$Age^2
# Split: 80% train / 20% test
split <- createDataPartition(df_model$Total_Purchase_Amount, p = 0.8, list = FALSE)
train_data <- df_model[split, ]
test_data <- df_model[-split, ]

```


```{r}
# Normalize numeric variables for linear models

pre_proc <- preProcess(train_data[, c("Age", "Age2")], method = c("center", "scale"))
train_data_norm <- train_data
test_data_norm <- test_data

train_data_norm[, c("Age", "Age2")] <- predict(pre_proc, train_data[, c("Age", "Age2")])
test_data_norm[, c("Age", "Age2")] <- predict(pre_proc, test_data[, c("Age", "Age2")])
```

##### 1. Linear Regression

```{r}
train_control <- trainControl(method = "cv", number = 5)
model_lm <- train(Total_Purchase_Amount ~ Age + Product_Category + Payment_Method,
                  data = train_data_norm, method = "lm", trControl = train_control)
```

##### 2. Polynomial Regression

```{r}
#model_poly <- train(Total_Purchase_Amount ~ Age + Age2 + Product_Category + Payment_Method,
                    #data = train_data_norm, method = "lm", trControl = train_control)
```

##### 3. Random Forest
```{r}
tune_grid <- expand.grid(mtry = c(2, 3))

model_rf <- train(
  Total_Purchase_Amount ~ Age + Product_Category + Payment_Method,
  data = train_data,
  method = "rf",
  trControl = trainControl(method = "cv", number = 3),
  tuneGrid = tune_grid,
  ntree = 100
)
```


```{r}
#model_rf <- train(Total_Purchase_Amount ~ Age + Product_Category + Payment_Method,
                  #data = train_data,
                  #method = "rf",
                  #trControl = train_control)
```

##### 4. Gradient Boosting (GBM)
```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("gbm")
library(gbm)
model_gbm <- train(Total_Purchase_Amount ~ Age + Product_Category + Payment_Method,
                   data = train_data,
                   method = "gbm",
                   trControl = train_control,
                   verbose = FALSE)
```


##### Summary Table

```{r}
pred_lm   <- predict(model_lm, test_data_norm)
#pred_poly <- predict(model_poly, test_data_norm)
pred_rf   <- predict(model_rf, test_data)
pred_gbm  <- predict(model_gbm, test_data)
```

```{r}
install.packages("Metrics")
library(Metrics)
library(caret)

results <- data.frame(
  Model = c("Linear Regression",  "Random Forest", "Gradient Boosting"),
  RMSE = c(
    rmse(test_data$Total_Purchase_Amount, pred_lm),
    rmse(test_data$Total_Purchase_Amount, pred_rf),
    rmse(test_data$Total_Purchase_Amount, pred_gbm)
  ),
  MAE = c(
    mae(test_data$Total_Purchase_Amount, pred_lm),
    mae(test_data$Total_Purchase_Amount, pred_rf),
    mae(test_data$Total_Purchase_Amount, pred_gbm)
  ),
  R2 = c(
    R2(pred_lm, test_data$Total_Purchase_Amount),
    R2(pred_rf, test_data$Total_Purchase_Amount),
    R2(pred_gbm, test_data$Total_Purchase_Amount)
  ),
  MAPE = c(
    mape(test_data$Total_Purchase_Amount, pred_lm),
    mape(test_data$Total_Purchase_Amount, pred_rf),
    mape(test_data$Total_Purchase_Amount, pred_gbm)
  )
)

# Print the results
print(results)
```


#### 3. Data Modeling - Predictive Analysis

```{r}
# Using product category as the only independent variable and 'Returns' as the target

# Load libraries
library(ggplot2)
library(dplyr)
library(caret)
library(randomForest)
library(pROC)
library(xgboost)
library(e1071)
```

```{r}
# Load and Prepare Dataset
df <- read.csv("ecommerce_customer_data_large.csv")

# Change the colnames with underscore instead of dot
colnames(df) <- gsub("\\.", "_", colnames(df))

# Fill missing Returns with 0
df$Returns[is.na(df$Returns)] <- 0
df$Returns <- factor(df$Returns, levels = c(0, 1))  # Ensure correct factor levels

# Convert to factor
df$Product_Category <- as.factor(df$Product_Category)

# Encode numeric for XGBoost
df$Product_Category_Num <- as.numeric(df$Product_Category)
```

```{r}
# Train-test split (80-20)
set.seed(42)
train_index <- createDataPartition(df$Returns, p = 0.8, list = FALSE)
train_data <- df[train_index, ]
test_data <- df[-train_index, ]
```

```{r}
# Initialize result data frame
df_metrics <- data.frame(
  Model = character(),
  Accuracy = numeric(),
  Precision = numeric(),
  Recall = numeric(),
  F1_Score = numeric(),
  stringsAsFactors = FALSE
)
```


##### 1. Logistic Regression

```{r}
log_model <- glm(Returns ~ Product_Category, data = train_data, family = "binomial")
log_probs <- predict(log_model, newdata = test_data, type = "response")
log_preds <- ifelse(log_probs > 0.5, 1, 0)
cm_log <- confusionMatrix(factor(log_preds), factor(test_data$Returns), positive = "1")

df_metrics <- rbind(df_metrics, data.frame(
  Model = "Logistic Regression",
  Accuracy = cm_log$overall["Accuracy"],
  Precision = cm_log$byClass["Precision"],
  Recall = cm_log$byClass["Recall"],
  F1_Score = cm_log$byClass["F1"]
))

roc_log <- roc(test_data$Returns, log_probs)
plot(roc_log, main = "ROC Curve Comparison")
```


##### 2. Random Forest

```{r}
rf_model <- randomForest(Returns ~ Product_Category, data = train_data, ntree = 100)
rf_probs <- predict(rf_model, newdata = test_data, type = "prob")[,2]
rf_preds <- ifelse(rf_probs > 0.5, 1, 0)
cm_rf <- confusionMatrix(factor(rf_preds), factor(test_data$Returns), positive = "1")

df_metrics <- rbind(df_metrics, data.frame(
  Model = "Random Forest",
  Accuracy = cm_rf$overall["Accuracy"],
  Precision = cm_rf$byClass["Precision"],
  Recall = cm_rf$byClass["Recall"],
  F1_Score = cm_rf$byClass["F1"]
))

roc_rf <- roc(test_data$Returns, rf_probs)
plot(roc_rf, col = "blue", main = "ROC Curve - Random Forest")
```

##### 3. XGBoost

```{r}
# Convert Returns to numeric for XGBoost (ensure 0 and 1)
train_label <- as.numeric(as.character(train_data$Returns))
test_label <- as.numeric(as.character(test_data$Returns))

# DMatrix for XGBoost
train_matrix <- xgb.DMatrix(data = as.matrix(train_data$Product_Category_Num), label = train_label)
test_matrix <- xgb.DMatrix(data = as.matrix(test_data$Product_Category_Num), label = test_label)

# Train XGBoost model
xgb_model <- xgboost(data = train_matrix, objective = "binary:logistic", nrounds = 50, verbose = 0)

# Predict and evaluate
xgb_probs <- predict(xgb_model, test_matrix)
xgb_preds <- ifelse(xgb_probs > 0.5, 1, 0)
cm_xgb <- confusionMatrix(factor(xgb_preds), factor(test_data$Returns), positive = "1")

# Add to df_metrics
df_metrics <- rbind(df_metrics, data.frame(
  Model = "XGBoost",
  Accuracy = cm_xgb$overall["Accuracy"],
  Precision = cm_xgb$byClass["Precision"],
  Recall = cm_xgb$byClass["Recall"],
  F1_Score = cm_xgb$byClass["F1"]
))

roc_xgb <- roc(test_data$Returns, xgb_probs)
plot(roc_xgb, col = "darkgreen", main = "ROC Curve - XGBoost")

print(df_metrics)
```

```{r}
# Checking why the result are the same for 3 models
table(df$Product_Category, df$Returns)
```

- Analysis showed that return rates across product categories are remarkably similar (~40%), indicating that Product_Category alone is not a strong predictor of return behavior. Recommend to incorporate additional customer, transactional, or product-level features to build a more predictive model.

- Logistic Regression model have poor performance, recommend to use only tree-based models such as random forest and XGBoost.


##### Feature Importance

```{r}
# Using only 4 feature importance (Total_Purchase_Amount, Product_Price, Age, Purchase_Month)

# Load required libraries (4 important features without hyperparameter tuning)
library(caret)
library(dplyr)
library(xgboost)
library(randomForest)
library(lubridate)

# Data Preprocessing

# Load your dataset
df <- read.csv("ecommerce_customer_data_large.csv")

# Replace dots in column names
colnames(df) <- gsub("\\.", "_", colnames(df))

# Fill missing Returns with 0 and convert to integer
df$Returns[is.na(df$Returns)] <- 0
df$Returns <- as.integer(df$Returns)

# Convert Purchase_Date and engineer features
df$Purchase_Date <- as.Date(df$Purchase_Date, format = "%Y-%m-%d")
df$Purchase_Month <- month(df$Purchase_Date)
df$Purchase_Year <- format(df$Purchase_Date, "%Y")

# Drop Customer_Age
df <- df %>% dplyr::select(-Customer_Age)

# Select only required features
df_model <- df %>%
  dplyr::select(Returns, Total_Purchase_Amount, Product_Price, Age, Purchase_Month)

# Train-test split
set.seed(123)
train_index <- createDataPartition(df_model$Returns, p = 0.8, list = FALSE)
train_data <- df_model[train_index, ]
test_data  <- df_model[-train_index, ]

# True labels for evaluation
true_labels <- factor(test_data$Returns, levels = c(0, 1))

# Random Forest
#set.seed(123)
#rf_model <- randomForest(as.factor(Returns) ~ ., data = train_data)
#rf_preds <- predict(rf_model, test_data[, -1])
set.seed(123)
train_data_small <- train_data %>% dplyr::sample_frac(0.3)  # use 30% of training data
rf_model <- randomForest(as.factor(Returns) ~ ., data = train_data_small)

# XGBoost
train_matrix <- xgb.DMatrix(data = as.matrix(train_data[, -1]), label = train_data$Returns)
test_matrix <- xgb.DMatrix(data = as.matrix(test_data[, -1]), label = test_data$Returns)

xgb_model <- xgboost(
  data = train_matrix,
  objective = "binary:logistic",
  nrounds = 100,
  max_depth = 6,
  eta = 0.1,
  verbose = 0
)

xgb_probs <- predict(xgb_model, test_matrix)
xgb_preds <- ifelse(xgb_probs > 0.5, 1, 0)
```

```{r}
length(rf_preds)
length(xgb_preds)
length(true_labels)
```


```{r}
evaluate_model <- function(preds, true_labels, model_name) {
  preds <- factor(preds, levels = c(0, 1))
  true_labels <- factor(true_labels, levels = c(0, 1))  # add this to avoid level mismatch
  cm <- confusionMatrix(preds, true_labels, positive = "1")
  data.frame(
    Model = model_name,
    Accuracy = round(cm$overall["Accuracy"], 4),
    Precision = round(cm$byClass["Precision"], 4),
    Recall = round(cm$byClass["Recall"], 4),
    F1_Score = round(cm$byClass["F1"], 4)
  )
}
```

```{r}
# Trim true_labels and xgb_preds to match rf_preds
true_labels_trimmed <- true_labels[1:length(rf_preds)]
xgb_preds_trimmed <- xgb_preds[1:length(rf_preds)]

# Now evaluate
results <- bind_rows(
  evaluate_model(rf_preds, true_labels_trimmed, "Random Forest"),
  evaluate_model(xgb_preds_trimmed, true_labels_trimmed, "XGBoost")
)
```

```{r}
# Print results
print(results)
```
```{r}
# Feature Importance (random forest)
varImpPlot(rf_model)

# Feature Importance (XGBoost)
importance <- xgb.importance(model = xgb_model)
xgb.plot.importance(importance)
```

From both feature importance (from Random Forest and XGBoost), we can observed that Total_Purchase_Amount, Product_Price, Age, Purchase_Month are the top 4 feature importance to be used for predictive analysis.


##### Top 4 Feature Importance vs Returns

```{r}
# Using only 4 feature importance (Total_Purchase_Amount, Product_Price, Age, Purchase_Month)

# Load required libraries (4 important features without hyperparameter tuning)
library(caret)
library(dplyr)
library(xgboost)
library(randomForest)
library(lubridate)

# Data Preprocessing

# Load your dataset
df <- read.csv("ecommerce_customer_data_large.csv")

# Replace dots in column names
colnames(df) <- gsub("\\.", "_", colnames(df))

# Fill missing Returns with 0 and convert to integer
df$Returns[is.na(df$Returns)] <- 0
df$Returns <- as.integer(df$Returns)

# Convert Purchase_Date and engineer features
df$Purchase_Date <- as.Date(df$Purchase_Date, format = "%Y-%m-%d")
df$Purchase_Month <- month(df$Purchase_Date)
df$Purchase_Year <- format(df$Purchase_Date, "%Y")

# Drop Customer_Age
df <- df %>% dplyr::select(-Customer_Age)

# Select only required features
df_model <- df %>%
  dplyr::select(Returns, Total_Purchase_Amount, Product_Price, Age, Purchase_Month)

# Train-test split
set.seed(123)
train_index <- createDataPartition(df_model$Returns, p = 0.8, list = FALSE)
train_data <- df_model[train_index, ]
test_data  <- df_model[-train_index, ]

# True labels for evaluation
true_labels <- factor(test_data$Returns, levels = c(0, 1))

# Random Forest
#set.seed(123)
#rf_model <- randomForest(as.factor(Returns) ~ ., data = train_data)
#rf_preds <- predict(rf_model, test_data[, -1])
set.seed(123)
train_sample <- train_data %>% dplyr::sample_frac(0.3)  # Use 30% of data
rf_model <- randomForest(as.factor(Returns) ~ ., data = train_sample, ntree = 100)

# XGBoost
train_matrix <- xgb.DMatrix(data = as.matrix(train_data[, -1]), label = train_data$Returns)
test_matrix <- xgb.DMatrix(data = as.matrix(test_data[, -1]), label = test_data$Returns)

xgb_model <- xgboost(
  data = train_matrix,
  objective = "binary:logistic",
  nrounds = 100,
  max_depth = 6,
  eta = 0.1,
  verbose = 0
)

xgb_probs <- predict(xgb_model, test_matrix)
xgb_preds <- ifelse(xgb_probs > 0.5, 1, 0)

# Evaluation Function
evaluate_model <- function(preds, true_labels, model_name) {
  preds <- factor(preds, levels = c(0, 1))
  cm <- confusionMatrix(preds, true_labels, positive = "1")
  data.frame(
    Model = model_name,
    Accuracy = round(cm$overall["Accuracy"], 4),
    Precision = round(cm$byClass["Precision"], 4),
    Recall = round(cm$byClass["Recall"], 4),
    F1_Score = round(cm$byClass["F1"], 4)
  )
}
```


```{r}
# Ensure alignment
min_len <- min(length(rf_preds), length(xgb_preds), length(true_labels))
rf_preds_trimmed <- rf_preds[1:min_len]
xgb_preds_trimmed <- xgb_preds[1:min_len]
true_labels_trimmed <- true_labels[1:min_len]

# Evaluation function (define above this chunk)
evaluate_model <- function(preds, true_labels, model_name) {
  preds <- factor(preds, levels = c(0, 1))
  true_labels <- factor(true_labels, levels = c(0, 1))
  
  cm <- confusionMatrix(preds, true_labels, positive = "1")
  
  data.frame(
    Model = model_name,
    Accuracy = round(cm$overall["Accuracy"], 4),
    Precision = round(cm$byClass["Precision"], 4),
    Recall = round(cm$byClass["Recall"], 4),
    F1_Score = round(cm$byClass["F1"], 4)
  )
}

# Evaluate all models using trimmed inputs
results <- bind_rows(
  evaluate_model(rf_preds_trimmed, true_labels_trimmed, "Random Forest"),
  evaluate_model(xgb_preds_trimmed, true_labels_trimmed, "XGBoost")
)

# Output
print(results)

```

```{r}
# Check class ratio

sum(df$Returns == 1) / sum(df$Returns == 0)
```

The class ratio is 0.683. It means that for every 1 negative (non-return), there are approximately 0.68 positive cases (returns).

Returns (1) ≈ 40.6%

No Returns (0) ≈ 59.4%


##### Class weight

Class weights are used to assign different importance (penalty) to each class when training a model. This helps the model pay more attention to the minority class during training.

In randomForest(), classwt = c("0" = 1, "1" = 1.5) means the model will penalize misclassifying a return (1) more than misclassifying a non-return (0).

In xgboost, we use weight = ifelse(Returns == 1, 1.5, 1) to give higher weight to return instances.

```{r}
# Load required libraries (using calss weight to improve the performance)
library(caret)
library(dplyr)
library(xgboost)
library(randomForest)
library(lubridate)



# Load your dataset
df <- read.csv("ecommerce_customer_data_large.csv")

# Replace dots in column names
colnames(df) <- gsub("\\.", "_", colnames(df))

# Fill missing Returns with 0 and convert to integer
df$Returns[is.na(df$Returns)] <- 0
df$Returns <- as.integer(df$Returns)

# Convert Purchase_Date and engineer features
df$Purchase_Date <- as.Date(df$Purchase_Date, format = "%Y-%m-%d")
df$Purchase_Month <- month(df$Purchase_Date)
df$Purchase_Year <- format(df$Purchase_Date, "%Y")

# Drop Customer_Age
df <- df %>% dplyr::select(-Customer_Age)

# Select only required features
df_model <- df %>%
  dplyr::select(Returns, Total_Purchase_Amount, Product_Price, Age, Purchase_Month)

# Train-test split
set.seed(123)
train_index <- createDataPartition(df_model$Returns, p = 0.8, list = FALSE)
train_data <- df_model[train_index, ]
test_data  <- df_model[-train_index, ]

# True labels for evaluation
true_labels <- factor(test_data$Returns, levels = c(0, 1))

# Random Forest
set.seed(123)
rf_model <- randomForest(
  as.factor(Returns) ~ .,
  data = train_data,
  classwt = c("0" = 1, "1" = 1.5),  # class weight to penalize minority class
  ntree = 100
)
rf_preds <- predict(rf_model, test_data[, -1])

# XGBoost with class weights
train_matrix <- xgb.DMatrix(
  data = as.matrix(train_data[, -1]),
  label = train_data$Returns,
  weight = ifelse(train_data$Returns == 1, 1.5, 1)
)
test_matrix <- xgb.DMatrix(data = as.matrix(test_data[, -1]), label = test_data$Returns)

xgb_model <- xgboost(
  data = train_matrix,
  objective = "binary:logistic",
  nrounds = 100,
  max_depth = 6,
  eta = 0.1,
  verbose = 0
)
xgb_probs <- predict(xgb_model, test_matrix)
xgb_preds <- ifelse(xgb_probs > 0.5, 1, 0)

# Evaluation Function
evaluate_model <- function(preds, true_labels, model_name) {
  preds <- factor(preds, levels = c(0, 1))
  cm <- confusionMatrix(preds, true_labels, positive = "1")
  data.frame(
    Model = model_name,
    Accuracy = round(cm$overall["Accuracy"], 4),
    Precision = round(cm$byClass["Precision"], 4),
    Recall = round(cm$byClass["Recall"], 4),
    F1_Score = round(cm$byClass["F1"], 4)
  )
}

# Evaluate all models
results <- bind_rows(
  evaluate_model(rf_preds, true_labels, "Random Forest"),
  evaluate_model(xgb_preds, true_labels, "XGBoost")
)

# Print results
print(results)
```

### Conclusion

This analysis applied classification, regression, and predictive techniques to understand customer behavior in three key areas: churn, returns, and spending.

Churn Prediction: XGBoost combined with SMOTE achieved the best results, with exceptionally high precision and improved recall. This model enables businesses to proactively identify customers at risk of leaving and take targeted retention actions.

Return Prediction: Although return rates were consistent (~40%) across product categories, category alone was not a strong predictor. The most effective models used top features beyond category, offering deeper insights into return behavior and guiding inventory and policy decisions.

Spending Prediction: Regression models struggled to accurately predict customer spend, showing very low R² values. The random residuals suggest predictions were unbiased but lacked strong explanatory variables.

Overall, while classification tasks like churn detection performed well, regression and category-based predictions highlight the need for richer behavioral and product-level features. Future improvements should focus on advanced feature engineering to enhance model accuracy and business value.
